Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2018-03-12 paper/main.tex
Added 'revisions-subtractive.txt', updated with the remaining modifications needed to complete the paper. Applied some modifications to the text, rewriting the motivation section

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view, and perhaps even partially confirmed some of Chomsky's fears. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation. We illustrate how this can be done by simply exploiting Google Translate to map sentences from a gender neutral language into English. As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that senteces about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.

@@

If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights. One could in principle group different professional occupations according to their broad category (artistic, scientific, industrial, etc.) and evaluate whether the proportion of male pronouns is more pronounced in some fields compared to others. We expect, for example, that most sentences about computer-related jobs will be translated with male pronouns, as a reflection of the gender asymmetry of this job market. We can also collect translation statistics among varied gender neutral languages to see whether different cultures make different assumptions about the gender of a subject.

If conclusive, in our view such an analysis would suggest that automatic translation tools should be designed in such a way as to avoid becoming hostages to their own underlying statistical models. Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed. This poses 
mostly an ethical rather than technical limitation to statistical machine translation and big data statistical models in general.

@@

We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate. The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows. Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}). Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they
( 1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.

@@

\section{Conclusions}
In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them. As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society. By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process. Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis yet. Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones). We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese. Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.

We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.

>>>>

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation. We illustrate how this can be done by simply exploiting Google Translate to map sentences from a gender neutral language into English. As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that sentences about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.

@@

As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}. Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}. Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}. This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}. Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure. The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist. This paper intends to investigate whether it does. We are optimistic that our research endeavors, if conclusive, can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.

@@

We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate. The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows. Given the success of Google Translate, which amounts to 200 million users daily \cite{Gtranslate200daily2017}, we have chosen to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}). Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they
( 1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.

@@

\section{Conclusions}
In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them. As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society. By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process. Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings about gender bias in professional occupations possibly generalize to most translations from gender neutral idioms in the same tool, although we have not tested this hypothesis yet. Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones). We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese. Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.

We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.
